---
title: "EDA, Scraping, and Methodology STA160 Final"
output:
  html_document: default
date: "2024-06-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Data
```{r}
library(baseballr)
library(dplyr)
```


```{r}
library(ggplot2)
library(reshape2)
library(zoo)
```
```{r}
player_id <- playerid_lookup("Kershaw") %>%
  dplyr::filter(first_name == "Clayton") %>%
  dplyr::select(mlbam_id, first_name, last_name)
player_id[1,1]

# 477132
```
```{r}
# updated functions in the package `baseballr` 

csv_from_url <- function(...){
  data.table::fread(...)
}
#-------------------------------------------------------------------------------------
make_baseballr_data <- function(df, type, timestamp){
  out <- df %>%
    tidyr::as_tibble()

  class(out) <- c("baseballr_data","tbl_df","tbl","data.table","data.frame")
  attr(out,"baseballr_timestamp") <- timestamp
  attr(out,"baseballr_type") <- type
  return(out)
}
# @title
statcast_search <- function(start_date = Sys.Date() - 1, end_date = Sys.Date(),
                                   playerid = NULL,
                                   player_type = "batter", ...) {
  # Check for other user errors.
  if (start_date <= "2015-03-01") { # March 1, 2015 was the first date of Spring Training.
    message("Some metrics such as Exit Velocity and Batted Ball Events have only been compiled since 2015.")
  }
  if (start_date < "2008-03-25") { # March 25, 2008 was the first date of the 2008 season.
    stop("The data are limited to the 2008 MLB season and after.")
    return(NULL)
  }
  if (start_date == Sys.Date()) {
    message("The data are collected daily at 3 a.m. Some of today's games may not be included.")
  }
  if (start_date > as.Date(end_date)) {
    stop("The start date is later than the end date.")
    return(NULL)
  }

  playerid_var <- ifelse(player_type == "pitcher",
                         "pitchers_lookup%5B%5D", "batters_lookup%5B%5D")

  vars <- tibble::tribble(
    ~var, ~value,
    "all", "true",
    "hfPT", "",
    "hfAB", "",
    "hfBBT", "",
    "hfPR", "",
    "hfZ", "",
    "stadium", "",
    "hfBBL", "",
    "hfNewZones", "",
    "hfGT", "R%7CPO%7CS%7C&hfC",
    "hfSea", paste0(lubridate::year(start_date), "%7C"),
    "hfSit", "",
    "hfOuts", "",
    "opponent", "",
    "pitcher_throws", "",
    "batter_stands", "",
    "hfSA", "",
    "player_type", player_type,
    "hfInfield", "",
    "team", "",
    "position", "",
    "hfOutfield", "",
    "hfRO", "",
    "home_road", "",
    playerid_var, ifelse(is.null(playerid), "", as.character(playerid)),
    "game_date_gt", as.character(start_date),
    "game_date_lt", as.character(end_date),
    "hfFlag", "",
    "hfPull", "",
    "metric_1", "",
    "hfInn", "",
    "min_pitches", "0",
    "min_results", "0",
    "group_by", "name",
    "sort_col", "pitches",
    "player_event_sort", "h_launch_speed",
    "sort_order", "desc",
    "min_abs", "0",
    "type", "details") %>%
    dplyr::mutate(pairs = paste0(.data$var, "=", .data$value))

  if (is.null(playerid)) {
    # message("No playerid specified. Collecting data for all batters/pitchers.")
    vars <- vars %>%
      dplyr::filter(!grepl("lookup", .data$var))
  }

  url_vars <- paste0(vars$pairs, collapse = "&")
  url <- paste0("https://baseballsavant.mlb.com/statcast_search/csv?", url_vars)
  # message(url)

  # Do a try/catch to show errors that the user may encounter while downloading.
  tryCatch(
    {
      suppressMessages(
        suppressWarnings(
          payload <- csv_from_url(url, encoding ="UTF-8")
        )
      )
    },
    error = function(cond) {
      message(cond)
      stop("No payload acquired")
    },
    # this will never run??
    warning = function(cond) {
      message(cond)
    }
  )
  # returns 0 rows on failure but > 1 columns
  if (nrow(payload) > 1) {

    names(payload) <- c("pitch_type", "game_date", "release_speed", "release_pos_x",
                        "release_pos_z", "player_name", "batter", "pitcher", "events",
                        "description", "spin_dir", "spin_rate_deprecated", "break_angle_deprecated",
                        "break_length_deprecated", "zone", "des", "game_type", "stand",
                        "p_throws", "home_team", "away_team", "type", "hit_location",
                        "bb_type", "balls", "strikes", "game_year", "pfx_x", "pfx_z",
                        "plate_x", "plate_z", "on_3b", "on_2b", "on_1b", "outs_when_up",
                        "inning", "inning_topbot", "hc_x", "hc_y", "tfs_deprecated",
                        "tfs_zulu_deprecated", "fielder_2", "umpire", "sv_id", "vx0",
                        "vy0", "vz0", "ax", "ay", "az", "sz_top", "sz_bot", "hit_distance_sc",
                        "launch_speed", "launch_angle", "effective_speed", "release_spin_rate",
                        "release_extension", "game_pk", "pitcher_1", "fielder_2_1",
                        "fielder_3", "fielder_4", "fielder_5", "fielder_6", "fielder_7",
                        "fielder_8", "fielder_9", "release_pos_y", "estimated_ba_using_speedangle",
                        "estimated_woba_using_speedangle", "woba_value", "woba_denom",
                        "babip_value", "iso_value", "launch_speed_angle", "at_bat_number",
                        "pitch_number", "pitch_name", "home_score", "away_score", "bat_score",
                        "fld_score", "post_away_score", "post_home_score", "post_bat_score",
                        "post_fld_score", "if_fielding_alignment", "of_fielding_alignment",
                        "spin_axis", "delta_home_win_exp", "delta_run_exp", "bat_speed", "swing_length")
    payload <- process_statcast_payload(payload) %>%
      make_baseballr_data("MLB Baseball Savant Statcast Search data from baseballsavant.mlb.com",Sys.time())
    return(payload)
  } else {
    warning("No valid data found")


# (somewhere within the statcast_search function before the payload is searched for)
colos <- c("pitch_type", "game_date",
            "release_speed", "release_pos_x", "release_pos_z",
            "player_name", "batter", "pitcher",
            "events", "description", "spin_dir",
            "spin_rate_deprecated", "break_angle_deprecated",
            "break_length_deprecated", "zone", "des",
            "game_type", "stand", "p_throws",
            "home_team", "away_team", "type",
            "hit_location", "bb_type", "balls",
            "strikes", "game_year", "pfx_x",
            "pfx_z", "plate_x", "plate_z",
            "on_3b", "on_2b", "on_1b", "outs_when_up",
            "inning", "inning_topbot", "hc_x",
            "hc_y", "tfs_deprecated", "tfs_zulu_deprecated",
            "fielder_2", "umpire", "sv_id",
            "vx0", "vy0", "vz0", "ax",
            "ay", "az", "sz_top", "sz_bot",
            "hit_distance_sc", "launch_speed", "launch_angle",
            "effective_speed", "release_spin_rate",
            "release_extension", "game_pk", "pitcher_1",
            "fielder_2_1", "fielder_3", "fielder_4",
            "fielder_5", "fielder_6", "fielder_7",
            "fielder_8", "fielder_9", "release_pos_y",
            "estimated_ba_using_speedangle", "estimated_woba_using_speedangle",
            "woba_value", "woba_denom", "babip_value",
            "iso_value", "launch_speed_angle", "at_bat_number",
            "pitch_number", "pitch_name", "home_score",
            "away_score", "bat_score", "fld_score",
            "post_away_score", "post_home_score",
            "post_bat_score", "post_fld_score", "if_fielding_alignment",
            "of_fielding_alignment", "spin_axis",
            "delta_home_win_exp", "delta_run_exp")
colNumber <- ncol(payload)
if(length(colos) != colNumber){
  newCols <- paste("newStat", 1:(length(colos) - colNumber))
  colos <- c(colos, newCols)
  message("New stats detected! baseballr will be updated soon to properly identify these stats")
}
# payload is acquired somewhere in here
# when the payload columns need to be named:
names(payload) <- colos



payload <- payload %>%
  make_baseballr_data("MLB Baseball Savant Statcast Search data from baseballsavant.mlb.com",Sys.time())
    return(payload)
  }
}

statcast_search.default <- function(start_date = Sys.Date() - 1, end_date = Sys.Date(),
                                             playerid = NULL, player_type = "batter", ...) {

  message(paste0(start_date, " is not a date. Attempting to coerce..."))
  start_Date <- as.Date(start_date)

  tryCatch(
    {
      end_Date <- as.Date(end_date)
    },
    warning = function(cond) {
      message(paste0(end_date, " was not coercible into a date. Using today."))
      end_Date <- Sys.Date()
      message("Original warning message:")
      message(cond)
    }
  )

  statcast_search(start_Date, end_Date,
                         playerid, player_type, ...)

}

statcast_search_batters <- function(start_date, end_date, batterid = NULL, ...) {
  statcast_search(start_date, end_date, playerid = batterid,
                         player_type = "batter", ...)
}


statcast_search_pitchers <- function(start_date, end_date, pitcherid = NULL, ...) {
  statcast_search(start_date, end_date, playerid = pitcherid,
                  player_type = "pitcher", ...)
}
```


```{r}
player_15 <- statcast_search(start_date ="2015-03-31", end_date ="2015-10-31", playerid = 477132, player_type = 'pitcher')
player_16 <- statcast_search(start_date ="2016-03-31", end_date ="2016-10-31", playerid = 477132, player_type = 'pitcher')
player_17 <- statcast_search(start_date ="2017-03-31", end_date ="2017-10-31", playerid = 477132, player_type = 'pitcher')
player_18 <- statcast_search(start_date ="2018-03-31", end_date ="2018-10-31", playerid = 477132, player_type = 'pitcher')
player_19 <- statcast_search(start_date ="2019-03-31", end_date ="2019-10-31", playerid = 477132, player_type = 'pitcher')

# type  ball or strike or ball in play
player1 <- bind_rows(player_15,player_16,player_17,player_18,player_19) %>%
  mutate(Year = as.factor(substr(game_date,1,4))) %>%
  filter(type == "S") %>%
  filter(launch_speed != 0)

# type  strike
player2 <- bind_rows(player_15,player_16,player_17,player_18,player_19) %>%
  mutate(Year = as.factor(substr(game_date,1,4))) %>%
  filter(type == "S")

# type  ball in play only
player3 <- bind_rows(player_15,player_16,player_17,player_18,player_19) %>%
  mutate(Year = as.factor(substr(game_date,1,4))) %>%
  filter(launch_speed != 0)  

# any pitch
player4 <- bind_rows(player_15,player_16,player_17,player_18,player_19) %>%
  mutate(Year = as.factor(substr(game_date,1,4)))

# column names
print(colnames(player1))
```


```{r}
# Summary statistics for a release speed 
summary(player4$release_speed)

# Summary statistics for a pitch type 
table(player4$pitch_type) 
```


```{r}
boxplot(player1$release_speed, main="Boxplot of Release Speed", ylab="Release Speed", col="orange")

```
```{r}
barplot(table(player4$pitch_type), main="Pitch Type Distribution", xlab="Pitch Type", ylab="Frequency", col="purple", las=2)
```
```{r}
# Summary statistics for a release speed 
summary(player3$release_speed)

# Summary statistics for a pitch type 
table(player3$pitch_type)
```
```{r}
# Summary statistics for a release speed 
summary(player2$release_speed)

# Summary statistics for a pitch type 
table(player2$pitch_type)
```
Player3 represents ball that were in play. Player2 represents balls that were strikes. We can see that there is not a big difference in ball speed between the two outcomes.

```{r}
barplot(table(player3$pitch_type), main="Pitch Type Distribution", xlab="Pitch Type", ylab="Frequency", col="purple", las=2)
```
```{r}
barplot(table(player2$pitch_type), main="Pitch Type Distribution", xlab="Pitch Type", ylab="Frequency", col="purple", las=2)
```
The plots above show that the type of pitch is similar between type strikes and types ball in play. It seems that his favorite pitches to use are the four seem fastball, the slider, and the curve ball.

```{r}
library(ggplot2)

ggplot(player4, aes(x=release_speed)) + 
  geom_histogram(binwidth=1, fill="blue", color="black") + 
  facet_grid(pitch_type ~ game_year) + 
  labs(title="Release Speed Distribution by Pitch Type and Game Year", x="Release Speed", y="Count")
```
We notice that as we go from 2015 to 2019, we see slight decrease in the speed of the FF pitches.

```{r}
player_avg_yr <- player1 %>%
  group_by(Year) %>%
  summarise(
    angle = round(mean(launch_angle, na.rm = TRUE),1),
    speed = round(mean(launch_speed, na.rm = TRUE),1)) 

player_avg_yr
```
```{r}
player1 %>%
  ggplot(aes(game_date, launch_speed)) +
  geom_point(alpha = 0.5) +
  stat_smooth(aes(group = Year, color = Year),formula = y~x, method='loess')+
  ggtitle("Clayton Kershaw") +  
  ylab("Speed = MPH") +
  xlab("Date") 

```
We can see that in years 2015, 2016, and 2018, the launch speed from Kershaw's pitches slowly increase as the season passed. This could be to pitcher fatigue or weather. However, in year 2017, the launch speed first decreased and then increased at the end of the season. In 2019, it stayed remotely leveled throughout the season.

```{r}
player1 %>%
  filter(pitch_type=="FF") %>% 
  ggplot(aes(game_date, release_speed)) +
  geom_point(alpha = 0.5) +
  stat_smooth(aes(group = Year, color = Year),formula = y~x, method='loess')+
  ggtitle("Clayton Kershaw") +  
  ylab("Release Speed") +
  xlab("Date") 
```
Four-Seam Fastball is a pitch that is best if it is as fast as possible. Therefore, we can see a clear decline through the years of Kershaw's career. When it comes to each individual season, there is no clear pattern of it decreasing as the season goes on (which is surprising).

```{r}
player1 %>%
  filter(pitch_type=="SL") %>% 
  ggplot(aes(game_date, pfx_x)) +
  geom_point(alpha = 0.5) +
  stat_smooth(aes(group = Year, color = Year),formula = y~x, method='loess')+
  ggtitle("Clayton Kershaw") +  
  ylab("Horizontal Movement") +
  xlab("Date")
```
Slider pitches rely on horizontal movement to be effective. The higher the magnitude of movement, the better. We can see that the 2015 season had some of Kershaw's highest movement sliders but also some of his lowest, while the 2019 season maintained a leveled magnitude throughout the season. Different from his fast ball performances, there is no year by year trend of his performance decreasing when it comes to the slider.

```{r}
player1 %>%
  filter(pitch_type=="CU") %>% 
  ggplot(aes(game_date, pfx_z)) +
  geom_point(alpha = 0.5) +
  stat_smooth(aes(group = Year, color = Year),formula = y~x, method='loess')+
  ggtitle("Clayton Kershaw") +  
  ylab("Vertical Movement") +
  xlab("Date")
```
Curveball pitches rely on vertical movement to be effective. The higher the magnitude of movement, the better. We can see the level of vertical movement in his curveball pitches is highest in 2015, and it slowly begins to decrease from there. The 2017 season seemed the lowest level in this case. The level of decrease in performance in his curveball from 2016 to 2017 is surprising. This could be due to injury or etc.


# Methodology

### K-means clustering to calculate pitch effectiveness


```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(factoextra)

# Assuming player4 contains the complete data for the years 2015-2019
# Create the Year column
player4 <- player4 %>%
  mutate(Year = as.factor(format(as.Date(game_date), "%Y")))

# Select relevant features, including the 'events' column
pitch_data <- player4 %>%
  select(release_speed, release_spin_rate, pfx_x, pfx_z, launch_angle, launch_speed, Year, events) %>%
  na.omit()  # Remove rows with missing values

# Scale the data, excluding non-numeric columns
pitch_data_scaled <- scale(pitch_data %>% select(release_speed, release_spin_rate, pfx_x, pfx_z, launch_angle, launch_speed))

# Perform K-means clustering with the optimal number of clusters (3)
set.seed(123)
kmeans_result <- kmeans(pitch_data_scaled, centers = 3, nstart = 25)

# Add cluster assignment to the original data
pitch_data <- cbind(pitch_data, cluster = kmeans_result$cluster)

# Visualize the clusters
fviz_cluster(kmeans_result, data = pitch_data_scaled, geom = "point", ellipse.type = "convex")
```
By applying k-means clustering, we can classify all of Clayton Kershaw's pitches into one of three categories. Next, we'll analyze each cluster.

```{r}
# Create detailed profiles for each cluster
cluster_profiles <- pitch_data %>%
  group_by(cluster) %>%
  summarise(
    release_speed_mean = mean(release_speed),
    release_speed_sd = sd(release_speed),
    release_spin_rate_mean = mean(release_spin_rate),
    release_spin_rate_sd = sd(release_spin_rate),
    pfx_x_mean = mean(pfx_x),
    pfx_x_sd = sd(pfx_x),
    pfx_z_mean = mean(pfx_z),
    pfx_z_sd = sd(pfx_z),
    launch_angle_mean = mean(launch_angle),
    launch_angle_sd = sd(launch_angle),
    launch_speed_mean = mean(launch_speed),
    launch_speed_sd = sd(launch_speed)
  )

print(cluster_profiles)
```

Cluster 1 contains pitches with a moderate release speed and high spin rate. These pitches have a slight horizontal movement to the left (negative pfx_x) and a moderate vertical movement. Also, the launch angle is close to zero.

Cluster 2 consists of pitches with the highest release speed among the clusters, but with a relatively lower spin rate. These pitches have minimal horizontal movement and significant vertical movement. 

Cluster 3 represents pitches with the lowest release speed and moderate spin rate. These pitches exhibit slight horizontal movement to the left and a negative vertical movement, indicating a downward break (e.g., curveballs). Also, the mean launch angle is low.


```{r}
# Visualize the distribution of release speed for each cluster
ggplot(pitch_data, aes(x = release_speed, fill = factor(cluster))) +
  geom_density(alpha = 0.5) +
  labs(title = "Distribution of Release Speed by Cluster", x = "Release Speed (mph)", fill = "Cluster")
```

```{r}
# Plot pitch usage over time
pitch_data %>%
  group_by(Year, cluster) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = Year, y = count, color = factor(cluster), group = cluster)) +
  geom_line() +
  labs(title = "Pitch Usage Over Time", x = "Year", y = "Count", color = "Cluster")
```

```{r}
# Evaluate pitch effectiveness based on hits (single, double, triple, home_run)
effectiveness <- pitch_data %>%
  group_by(cluster) %>%
  summarise(
    hits = sum(events %in% c("single", "double", "triple", "home_run")),
    total_pitches = n(),
    hit_rate = hits / total_pitches
  )

print(effectiveness)

```
Lastly, we can calculate which types of pitches in Clayton Kershaw's arsenal are most effective by seeing how many times each pitch was hit. 


# Additional Visualizations
```{r}
# Hit rate by year and cluster
effectiveness_by_year <- pitch_data %>%
  group_by(Year, cluster) %>%
  summarise(
    hits = sum(events %in% c("single", "double", "triple", "home_run")),
    total_pitches = n(),
    hit_rate = hits / total_pitches
  )

ggplot(effectiveness_by_year, aes(x = Year, y = hit_rate, color = factor(cluster))) +
  geom_line() +
  geom_point() +
  labs(title = "Hit Rate by Year and Cluster", x = "Year", y = "Hit Rate", color = "Cluster")

```

```{r}
# Cluster composition over time
cluster_composition <- pitch_data %>%
  group_by(Year, cluster) %>%
  summarise(count = n()) %>%
  mutate(proportion = count / sum(count))

ggplot(cluster_composition, aes(x = Year, y = proportion, fill = factor(cluster))) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Cluster Composition Over Time", x = "Year", y = "Proportion", fill = "Cluster")

```

```{r}
# Scatter plot of release speed vs. launch angle colored by cluster
ggplot(pitch_data, aes(x = release_speed, y = launch_angle, color = factor(cluster))) +
  geom_point(alpha = 0.5) +
  labs(title = "Release Speed vs. Launch Angle by Cluster", x = "Release Speed (mph)", y = "Launch Angle", color = "Cluster")

```

```{r}

```
